---
title: "whale_data_exploration"
author: "Sarah Salois"
output: 
 html_document:
    code_fold: hide
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(wesanderson)
library(factoextra) # fviz_cluster
library(gridExtra)
library(cluster)
library(ggpubr)

```



```{r, data wrangling}
# color palette
pal = wes_palette("Darjeeling1")
# bring in data 
df <- read.csv('data/whales_rings_chl_tmp.csv')
df<- df %>% 
  mutate(date = lubridate::mdy(date),
         year = lubridate::year(date), 
         week = lubridate::week(date))
df.21.23 <- df %>% filter(year %in% c(2021:2023)) %>%
  mutate(num_ring = as.integer(num_ring))
# create a variable for proportion of rings associated with streamer
df.21.23$prop.streamering <- (df.21.23$num_ring_with_streamer)/(df.21.23$num_ring)
df_tally <- df %>%
  group_by(year) %>%
  summarise(ttl_right = sum(na.omit(right_whale_sight)), 
            ttl_other = sum(na.omit(other_whale_sight)), 
            ttl_unidentified = sum(na.omit(unidentified_whale_sight)))

df_tally$ttl_whale <- rowSums(df_tally[,c(2:4)])
tally.long <- df_tally %>%
  pivot_longer(ttl_right:ttl_whale, names_to = 'whales', values_to = 'count')



```

### Visualing Offshore Flight Summary Data

**Figure 1.** Total number of whale sightings from Offshore Flight data

Sightings were summed across years. Bars indicate the whale sightings, where red 
represents right whales, green represents other large whales and yellow indicates unidentified large 
whales.
```{r}
g = ggplot() +
  geom_col(data = tally.long %>% 
             filter(whales %in% c('ttl_right','ttl_other','ttl_unidentified')),
           aes(x = factor(year), y = count, fill = whales), 
           size = 1, color = 'black',
           stat='identity', position='dodge') +
  scale_fill_manual(values = pal,
                    name = 'Species',
                    labels = c('Right Whale',
                               'Other Large Whale',
                               'Unidentified Large Whale')) +
  # geom_line(data = tally.long %>% filter(whales =='ttl_whale'), 
  #                                 aes(x = factor(year), y = count), size = 1.5, 
  #           color = 'black', group = 1) +
  xlab('Year') +
  ylab('Number of Whale Sightings (n)') +
  #labs(title = 'Offshore Flight Data Summary')+
  theme_classic()  
g +  theme(text = element_text(size = 14), 
         axis.text = element_text(size = 13), 
         axis.title = element_text(size = 15),
         legend.position = c(0.25, 0.85),
         legend.background = element_rect(fill = "white", color = "black"), 
         legend.title = element_text(size = 11),
         legend.text = element_text(size = 11))
```


**Figure 2.** Number of whale sightings and number of rings with streamers 
over time. 

Bars indicate the whale sightings, where red represents right whales, 
green represents other large whales and yellow indicates unidentified large 
whales. The black line is the number of rings associated with a shelf streamer
present (West of -65W) on the date of sighting. Note, I did not have JC charts 
for all the years, thus the yellow points on top of the black line are showing 
the years for which I had JC charts to reference. This data is also based on a 
limited set of JC charts, with most years only having images from Jan-May.  

```{r}
g2 = ggplot() +
  geom_col(data = tally.long %>%
             filter(whales %in% c('ttl_right','ttl_other','ttl_unidentified')),
           aes(x = factor(year), y = count, fill = whales),
           size = 1, color = 'black',
           stat='identity', position='dodge') +
  scale_fill_manual(values = pal,
                    name = 'Species',
                    labels = c('Right Whale',
                               'Other Large Whale',
                               'Unidentified Large Whale')) +
  geom_line(data = tally.long %>% filter(whales =='ttl_streamer_ring'), 
            aes(x = factor(year), y = count), size = 1.5, 
            color = 'black', group = 1) +
  geom_point(data = tally.long %>% filter(whales =='ttl_streamer_ring' & year %in% c(2014, 2016, 2017, 2019, 2020, 2021, 2022,2023 )), 
             aes(x = factor(year), y = count), bg = 'yellow', size = 3, pch = 21,
             color = 'black', group = 1) +
  #scale_y_continuous(sec.axis = sec_axis(~., name = 'Catch (lbs)')) + 
  xlab('Year') +
  ylab('Number of Whale Sightings / Number of Rings w/Streamer') +
  #labs(title = 'Offshore Flight Data Summary')+
  theme_classic()  
g2 +  theme(text = element_text(size = 14), 
           axis.text = element_text(size = 13), 
           axis.title = element_text(size = 15),
           legend.position = c(0.25, 0.85),
           legend.background = element_rect(fill = "white", color = "black"), 
           legend.title = element_text(size = 11),
           legend.text = element_text(size = 11))

```


**Figure 3.** Number of whale sightings and percentage of rings with streamers 
over time. 

Bars indicate the whale sightings, where red represents right whales, 
green represents other large whales and yellow indicates unidentified large 
whales. The black line is the percentage of all rings present (West of -65) that 
were associated with a shelf streamer on the given date of sighting. Note, 
I did not have JC charts for all the years, thus the yellow points on top of 
the black line are showing the years for which I had JC charts to reference.
This data is also based on a limited set of JC charts, with most years only
having images from Jan-May. 

```{r}
g3 = ggplot() +
  geom_col(data = tally.long %>% 
             filter(whales %in% c('ttl_right','ttl_other','ttl_unidentified')),
           aes(x = factor(year), y = count, fill = whales), 
           size = 1, color = 'black',
           stat='identity', position='dodge') +
  scale_fill_manual(values = pal,
                    name = 'Species',
                    labels = c('Right Whale',
                               'Other Large Whale',
                               'Unidentified Large Whale')) +
  geom_line(data = tally.long %>% filter(whales =='prop_stream_ring'), 
            aes(x = factor(year), y = count*100), size = 1.5, 
            color = 'black', group = 1) +
  geom_point(data = tally.long %>% filter(whales =='prop_stream_ring' & year %in% c(2014, 2016, 2017, 2019, 2020, 2021, 2022,2023 )), 
             aes(x = factor(year), y = count*100), bg = 'yellow', size = 3, pch = 21,
             color = 'black', group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~., name = 'Proportion of rings w/streamer')) + 
  xlab('Year') +
  ylab('Number of Whale Sightings / Number of Rings w/Streamer') +
  #labs(title = 'Offshore Flight Data Summary')+
  theme_classic()  
g3 +  theme(text = element_text(size = 14), 
           axis.text = element_text(size = 13), 
           axis.title = element_text(size = 15),
           legend.position = c(0.25, 0.85),
           legend.background = element_rect(fill = "white", color = "black"), 
           legend.title = element_text(size = 11),
           legend.text = element_text(size = 11))
```




### Kmeans clustering 

K-means clustering (MacQueen 1967) is an unsupervised machine learning algorithm 
for partitioning a given data set into a set of k groups (i.e. k clusters). 
The value K (the number of groups) is specified ahead of time. The algorithm 
classifies objects in multiple groups (i.e., clusters), such that **objects within the same cluster are as similar as possible (i.e., high intra-class similarity)**, 
whereas objects from **different clusters are as dissimilar as possible (i.e., low inter-class similarity)**. In k-means clustering, each cluster is represented by
its center (i.e, centroid) which corresponds to the mean of points assigned to 
the cluster.

More details can be found [here](https://www.datanovia.com/en/lessons/k-means-clustering-in-r-algorith-and-practical-examples/)

#### Right Whales: 

```{r}
### Cluster : Right whales 
selected.vars <- df[, c('right_whale_sight', 
                              'num_ring_with_streamer',
                              'num_ring', 'prop.streamering')]
# remove NAs for kmeans to work 
selected.vars <- na.omit(selected.vars)
# optimal number of clusters:   
fviz_nbclust(selected.vars[,-c(3,4)], kmeans, method = "silhouette")

# -- Assigning a catagorical attribute to better understand clusters -- #
# Using proportion of rings that are associated with streamers
## High low ##
# selected.vars$prop <- case_when(
#   selected.vars$prop.streamering >= .5 ~ 'high',
#   selected.vars$prop.streamering < .5 ~ 'low'
# )
## High med low ##
selected.vars$prop <- case_when(
  selected.vars$prop.streamering >= .6 ~ 'high',
  selected.vars$prop.streamering >= .3 & selected.vars$prop.streamering < .6 ~ 'mod',
  selected.vars$prop.streamering <= .3 ~ 'low'
)
# Compute k-means with k = 2
set.seed(123)
res.km <- kmeans(scale(selected.vars[, -c(3,4,5)]), 2, nstart = 25)
# K-means clusters showing the group of each individuals
# res.km$cluster
# Dimension reduction using PCA
res.pca <- prcomp(selected.vars[,-c(3,4,5)],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(res.km$cluster)
# Add Species groups from the original data sett
ind.coord$prop <- selected.vars$prop
# Data inspection
head(ind.coord)

# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  shape = 'prop', size = 3,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)

```
 
#### Other Large Whales:  
 
```{r}

### Cluster : Other whales 
selected.vars.ow <- df[, c('other_whale_sight', 
                               'num_ring_with_streamer',
                               'num_ring', 'prop.streamering')]
selected.vars.ow <- na.omit(selected.vars.ow)
# optimal number of clusters:   
fviz_nbclust(selected.vars.ow[,-c(3,4)], kmeans, method = "silhouette")

# High low
selected.vars.ow$prop <- case_when(
  selected.vars.ow$prop.streamering >= .5 ~ 'high',
  selected.vars.ow$prop.streamering < .5 ~ 'low'
)
# High med low
selected.vars.ow$prop <- case_when(
  selected.vars.ow$prop.streamering >= .6 ~ 'high',
  selected.vars.ow$prop.streamering >= .3 & selected.vars.ow$prop.streamering < .6 ~ 'mod',
  selected.vars.ow$prop.streamering <= .3 ~ 'low'
)
# Compute k-means with k = 3
set.seed(123)
res.km <- kmeans(scale(selected.vars.ow[, -c(3,4,5)]), 3, nstart = 25)
# K-means clusters showing the group of each individuals
# res.km$cluster
# Dimension reduction using PCA
res.pca <- prcomp(selected.vars.ow[,-c(3,4,5)],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(res.km$cluster)
# Add Species groups from the original data sett
ind.coord$prop <- selected.vars.ow$prop
# Data inspection
head(ind.coord)

# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  shape = 'prop', size = 3,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

