---
title: "whale_data_exploration"
author: "Sarah Salois"
output: 
 html_document:
    code_fold: hide
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(wesanderson)
library(factoextra) # fviz_cluster
library(gridExtra)
library(cluster)
library(ggpubr)

```



```{r, data wrangling}
# color palette
pal = wes_palette("Darjeeling1")
# bring in data 
df <- read.csv('data/whales_rings_chl_tmp.csv')
df<- df %>% 
  mutate(date = lubridate::mdy(date),
         year = lubridate::year(date), 
         week = lubridate::week(date))
df.21.23 <- df %>% filter(year %in% c(2021:2023)) %>%
  mutate(num_ring = as.integer(num_ring))
# create a variable for proportion of rings associated with streamer
df.21.23$prop.streamering <- (df.21.23$num_ring_with_streamer)/(df.21.23$num_ring)
df_tally <- df %>%
  group_by(year) %>%
  summarise(ttl_right = sum(na.omit(right_whale_sight)), 
            ttl_other = sum(na.omit(other_whale_sight)), 
            ttl_unidentified = sum(na.omit(unidentified_whale_sight)))

df_tally$ttl_whale <- rowSums(df_tally[,c(2:4)])
tally.long <- df_tally %>%
  pivot_longer(ttl_right:ttl_whale, names_to = 'whales', values_to = 'count')



```

### Visualing Offshore Flight Summary Data

```{r}
g = ggplot() +
  geom_col(data = tally.long %>% 
             filter(whales %in% c('ttl_right','ttl_other','ttl_unidentified')),
           aes(x = factor(year), y = count, fill = whales), 
           size = 1, color = 'black',
           stat='identity', position='dodge') +
  scale_fill_manual(values = pal,
                    name = 'Species',
                    labels = c('Right Whale',
                               'Other Large Whale',
                               'Unidentified Large Whale')) +
  geom_line(data = tally.long %>% filter(whales =='ttl_whale'), 
                                  aes(x = factor(year), y = count), size = 1.5, 
            color = 'black', group = 1) +
  xlab('Year') +
  ylab('Number of Whale Sightings (n)') +
  #labs(title = 'Offshore Flight Data Summary')+
  theme_classic()  
g +  theme(text = element_text(size = 14), 
         axis.text = element_text(size = 13), 
         axis.title = element_text(size = 15),
         legend.position = c(0.25, 0.85),
         legend.background = element_rect(fill = "white", color = "black"), 
         legend.title = element_text(size = 11),
         legend.text = element_text(size = 11))
```


### Kmeans clustering 

K-means clustering (MacQueen 1967) is an unsupervised machine learning algorithm for partitioning a given data set into a set of k groups (i.e. k clusters). The analyst specifies k (the number of groups) pre-hoc. The algorithm classifies objects in multiple groups (i.e., clusters), such that **objects within the same cluster are as similar as possible (i.e., high intra-class similarity)**, whereas objects from **different clusters are as dissimilar as possible (i.e., low inter-class similarity)**. In k-means clustering, each cluster is represented by its center (i.e, centroid) which corresponds to the mean of points assigned to the cluster.

More details can be found [here](https://www.datanovia.com/en/lessons/k-means-clustering-in-r-algorith-and-practical-examples/)

#### Right Whales: 

```{r}
### Cluster : Right whales 
selected.vars3 <- df.21.23[, c('right_whale_sight', 
                              'num_ring_with_streamer',
                              'num_ring', 'prop.streamering')]
# remove NAs for kmeans to work 
selected.vars3 <- na.omit(selected.vars3)
# optimal number of clusters:   
fviz_nbclust(selected.vars3[,-c(3,4,5)], kmeans, method = "silhouette")

# -- Assigning a catagorical attribute to better understand clusters -- #
# Using proportion of rings that are associated with streamers
## High low ##
# selected.vars3$prop <- case_when(
#   selected.vars3$prop.streamering >= .5 ~ 'high',
#   selected.vars3$prop.streamering < .5 ~ 'low'
# )
## High med low ##
selected.vars3$prop <- case_when(
  selected.vars3$prop.streamering >= .6 ~ 'high',
  selected.vars3$prop.streamering >= .3 & selected.vars3$prop.streamering < .6 ~ 'mod',
  selected.vars3$prop.streamering <= .3 ~ 'low'
)
# Compute k-means with k = 2
set.seed(123)
res.km <- kmeans(scale(selected.vars3[, -c(3,4,5)]), 2, nstart = 25)
# K-means clusters showing the group of each individuals
# res.km$cluster
# Dimension reduction using PCA
res.pca <- prcomp(selected.vars3[,-c(4,5)],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(res.km$cluster)
# Add Species groups from the original data sett
ind.coord$prop <- selected.vars3$prop
# Data inspection
head(ind.coord)

# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  shape = 'prop', size = 3,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)

```
 
#### Other Large Whales:  
 
```{r}

### Cluster : Other whales 
selected.vars.ow <- df.21.23[, c('other_whale_sight', 
                               'num_ring_with_streamer',
                               'num_ring', 'prop.streamering')]
selected.vars.ow <- na.omit(selected.vars.ow)
# optimal number of clusters:   
fviz_nbclust(selected.vars.ow[,-c(3,4,5)], kmeans, method = "silhouette")

# High low
selected.vars.ow$prop <- case_when(
  selected.vars.ow$prop.streamering >= .5 ~ 'high',
  selected.vars.ow$prop.streamering < .5 ~ 'low'
)
# High med low
selected.vars.ow$prop <- case_when(
  selected.vars.ow$prop.streamering >= .6 ~ 'high',
  selected.vars.ow$prop.streamering >= .3 & selected.vars.ow$prop.streamering < .6 ~ 'mod',
  selected.vars.ow$prop.streamering <= .3 ~ 'low'
)
# Compute k-means with k = 2
set.seed(123)
res.km <- kmeans(scale(selected.vars.ow[, -c(3,4,5)]), 2, nstart = 25)
# K-means clusters showing the group of each individuals
# res.km$cluster
# Dimension reduction using PCA
res.pca <- prcomp(selected.vars.ow[,-c(4,5)],  scale = TRUE)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(res.km$cluster)
# Add Species groups from the original data sett
ind.coord$prop <- selected.vars.ow$prop
# Data inspection
head(ind.coord)

# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
head(eigenvalue)

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
  shape = 'prop', size = 3,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("Dim 1 (", variance.percent[1], "% )" ),
  ylab = paste0("Dim 2 (", variance.percent[2], "% )" )
) +
  stat_mean(aes(color = cluster), size = 4)
```

